---
title: "Using vecpack"
author: "Simon Barthelmé"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

\newcommand{\bm}[1]{\boldsymbol{\mathbf{#1}}}

```{r init,echo=FALSE,message=FALSE}
library(vecpack)
library(purrr)
```

vecpack solves a problem commonly encountered when working with optimisers (like optim) and MCMC samplers: they expect to work on a function f that takes a vector as input. This is fine when optimising over vectors, but sometimes f is really a function of a matrix and a vector, or a matrix and an image and a scalar, etc. You could concatenate all these arguments into a single vector, but that always involves lots of boilerplate and book-keeping, and it's easy to make errors. 
vecpack saves you from having to write most of the boilerplate. 

Packing a list of values into a single vector is easy enough:

```{r}
l <- list(a= matrix(1:4,2,2),b = 0.4)
sapply(l,as.vector) %>% do.call(c,.)
```

In vecpack, that functionality is provided by the vpack function:

```{r}
vpack(l)
```

Going the other way is more complicated: e.g., the vector 1,2,3,4 could be a 2x2 matrix  followed by a scalar, or one scalar followed by a 2x2 matrix, or two vectors of length 2 and 3, etc.
To invert the packing, we need to know the structure, and in vecpack the structure is inferred from an example:

```{r}
vunpack <- gen.vunpack(l)
vunpack(1:5)
vpack(l) %>% vunpack
```

# Using vecpack with optim

The first example is somewhat silly but hopefully you'll see where I'm going with this: let's say we're doing linear regression. 
The data are pairs (x,y), our model is 
$$ E(y) = a x + b $$
and the corresponding log-likelihood is "cfun", below: 

```{r}
x <- seq(0,1,l=20)
y <- 3*x - 2 +rnorm(length(x))
cfun <- function(a,b) sum((y-(a*x+b))^2)
```

We'll use optim and vecpack to optimise cfun: 

```{r}
guess <- list(a=0,b=1)
up <- gen.vunpack(guess) #Unpacking function
cfun.lifted <- lift_up(cfun,up) #Now cfun accepts a vector argument
opt <- optim(vpack(guess),cfun.lifted)
up(opt$par)
```

vecpack includes an interface to optim called vpoptim, which removes all of the boilerplate: just provide a cost function, a guess and you're set. 

```{r}
opt <- vpoptim(guess,cfun)
opt$par
```

Here's a better example: we seek to find a low-rank + diagonal decomposition of a matrix, i.e.

$$ \bm{X} \approx \bm{A}\bm{A}^t + \sigma^2 \bm{I} $$ 

```{r}
X <- cov(USArrests)
cfun <- function(A,sigma2)
{
    lrank <- tcrossprod(A,A) + sigma2*diag(nrow(A)) 
    sum((X-lrank)^2)
}
guess <- list(A=matrix(0,4,2),sigma2=1)
opt <- vpoptim(guess,cfun)
opt$par
```

## Which objects can vecpack pack? 

At the moment vecpack knows how to pack scalars, vectors, matrices, arrays and cimg objects (images from the [imager package](http://dahtah.github.io/imager)). 

## Computing numerical gradients

There's an interface to the function "grad" from the *numDeriv* package, that works along the same lines as vpoptim:

```{r}
#gradient of a function of two arguments
vpgrad(function(x,y) x+y,list(x=1,y=3))
#gradient of a matrix function, returned as a matrix
vpgrad(det,diag(2))
```

# Extending vecpack

vecpack is extensible: you can define your own ways of packing objects into vectors. For example, suppose we want to pack tri-diagonal matrices into vectors. All we need to do is create an S3 class for tridiagonal matrices, and define just three generics: as.vector and reshapefun. 

First, some utility functions to get and set values: 

```{r}
#Extract or set values above diagonal
utri <- function(M)
{
    M[row(M) == col(M) - 1]
}

`utri<-` <- function(M,value)
{
    M[row(M) == col(M) - 1] <- value
    M
}

#Extract or set values below
ltri <- function(M)
{
    M[row(M) == col(M) + 1]
}

`ltri<-` <- function(M,value)
{
    M[row(M) == col(M) + 1] <- value
    M
}
```




## First generic: as.vector 

as.vector should take a (tri-diagonal) matrix and pack it into a vector: 

```{r}
as.vector.trid <- function(M,...)
{
    c(ltri(M),diag(M),utri(M))
}

trid <- function(M)
{
    class(M) <- c('trid','matrix')
    M
}

diag(3) %>% trid %>% as.vector
```

## Second generic: length

length should return the length of the tridiagonal matrix *when packed as a vector*: a tridiagonal matrix has $n + 2*(n-1) = 3n - 2$ degrees of freedom.

```{r}
length.trid <- function(V)
{
    n <- nrow(V)
    3*n-2
}
```

## Third generic: reshapefun 

And next we need to define "reshapefun". Reshaping turns the vector back into the original object. "reshapefun.trid" should return a reshaping function appropriate for a given matrix. Here's reshapefun.matrix, for instance:

```{r}
reshapefun.matrix <- function(x,...) function(k) matrix(k,nrow(x),ncol(x))
rs <- matrix(0,2,2) %>% reshapefun
rs(1:4)
```

We define reshapefun.trid along the same lines:

```{r}
reshapefun.trid <- function(x,...)
{
    n <- nrow(x)
    #The following function returns a matrix of the right size
    function(k)
    {
        M <- matrix(0,n,n)
        ltri(M) <- k[1:(n-1)]
        diag(M) <- k[n+0:(n-1)]
        utri(M) <- k[(2*n):length(k)]
        M
    }
}
V <- diag(3) %>% trid
rs <- reshapefun(V)
as.vector(V) %>% rs
```

We're now good to go with vpack:

```{r}
l <- list(V=V,a=3)
up <- gen.vunpack(l)
vpack(l) %>% up
```

We can now use it to find, e.g., a tri-diagonal approximate inverse: 

```{r}
#The cost function encourages A*X = Id
X <- rnorm(16) %>% matrix(4,4)
cfun <- function(A)
{
    P <- A%*%X
    sum(abs(eigen(P)$val- 1))
}
A <- diag(4) %>% trid
opt <- vpoptim(A,cfun,method="BFGS",control=list(trace=T,maxit=1e5))
opt$par
```

## Making your own wrapper

If you'd like to wrap another optimiser/MCMC sampler, it should be easy enough. The source for vpoptim is just a few lines: 

```{r}
vpoptim <- function(guess,cfun,...)
{
    up <- gen.vunpack(guess) 
    cfun.lifted <- lift_up(cfun,up) 
    opt <- optim(vpack(guess),cfun.lifted,...)
    opt$par <- up(opt$par)
    opt
}
```

If you've made another interface, consider contributing it to the vecpack package. 

## Other uses for vecpack

You can use vecpack to express constraints as well (hiding them from the cost function). TODO: example with SPD matrices. 
